---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    fig_caption: yes
    keep_md: yes
---


## Loading and preprocessing the data
```{r csvread, echo=TRUE}
data <- read.csv("activity.csv", header=T)
```



## What is mean total number of steps taken per day?
so lets plot a histogram of the # of steps/day and take a summary:

```{r  meansteps, echo=TRUE}
agg_date <- aggregate(steps ~ date, data=data, FUN=sum)
hist(agg_date$steps, breaks=10, col="Red", main="", xlab="# steps/day")
summary(agg_date)
```
the summary reports mean = 10766 and median = 10765



## What is the average daily activity pattern?
```{r avgactivity, echo=TRUE}
agg_int <- aggregate(steps ~ interval, data=data, FUN=mean)
summary(agg_int)
plot(agg_int, type="l", main="Average Daily Activity Pattern", xlab="5-min interval", ylab="avg # steps/day")
```

The interval with the maximum # of steps was found by:
```{r maxsteps, echo=TRUE}
mx <- max(agg_int$steps)
agg_int[agg_int$steps==mx,]
```
indicating the average max # steps of 206.17 occurred at 8:35 am


## Imputing missing values
```{r datasummary, echo=TRUE}
summary(data)
```

From the summary above, we find 2,304 rows had NA entries for steps.

To fill these data holes, I used the mean # of steps/5-min interval (agg_int) data frame

Created a new column (wd_we) and converted the date to a weekday or weekend factor and saved it to a copy of the original data, called data_imp:

```{r dateconv, echo=TRUE}
#data_imp$day <- weekdays(as.Date(as.character(data_imp$date))) #tried this 1st; no workie
data_imp$wd_we <- as.factor(ifelse(!weekdays(as.Date(data_imp$date)) %in% c("Saturday", "Sunday"), "weekday","weekend"))
```

then I merged this new data frame with the agg_int data and filled all the NAs with the mean # of steps for that 5-min interval

```{r imputing, echo=TRUE}
df_merge <- merge(data_imp, agg_int, by.x="interval", by.y="interval")
df_merge$steps.x[is.na(df_merge$steps.x)] <- df_merge$steps.y
```

plotting a new histogram with the aggregated imputed data:
```{r histo2, echo=TRUE}
agg_imp <- aggregate(steps.x ~ date, data=df_merge, FUN=sum)
hist(agg_imp$steps.x, breaks=10, col="Red", main="", xlab="# steps/day")
summary(agg_imp)
```

From observations of both the histogram plot and the summary, it appears that imputing the NAs in the original data has lowered the mean and median values significantly.


## Are there differences in activity patterns between weekdays and weekends?
create a panel plot showing a time series for each set of data:

```{r panelplot, echo=TRUE}
library(lattice)
#xyplot(steps ~ interval | wd_we, data=data_imp, layout=c(2,1), type="l")
xyplot(steps.x ~ interval | wd_we, data=df_merge, layout=c(2,1), type="l")
```

Looking at the panel plots, there are distinct differences between the measurements made during the weekday and those of the weekend. Weekday step activity is significantly higher in the mornings after 5am, where weekend step activity doesn't really begin until after 8-9am.

The data also shows a common burst of activity that begins around 3pm for both data sets.
Both data sets also appear to have both an equal duration and pause between activity bursts. 



